The impact of forest management on the temperature sensitivity of organic matter turnover

### 2024-12-09 Manuscript_draft/Draft.Rmd
Remove all the management terms

# Introduction
The soil is one of the largest terrestrial reservoirs of carbon, containing approximately three times more carbon than the atmosphere or the vegetation on Earth (IPCC, 2021). Soil carbon is released to the atmosphere as carbon dioxide (CO2) through decomposition of organic matter by microbes (heterotrophic respiration) and root respiration (autotrophic respiration). This process (i.e., soil respiration) constitutes the main carbon efflux into the atmosphere (ref). Under the new climate change scenarios increases of CO2 efflux from soils are expected (Schlesinger and Andrews, 2000), this can imply a positive feedback to future warming if stored carbon is mobilized and carbon input to soil does not increase (Lamberty&Thomson, 2010). Because soil respiration is highly sensitive to soil temperature and soil moisture (Nissan et al. 2022), understanding the responses of soil respiration to these microclimatic conditions, which are affected by global change factors as climate change and forest management, is critical for predicting future atmospheric CO2 concentrations. Temperature affects respiration rates positively (Lloyd and Taylor, 1994), by increasing the metabolic rates of soil microorganisms and enzyme efficiency (ref). However, the relationship with soil moisture is non-linear. Low water content limit nutrient diffusion and enzymes and exert drought-stress on microbes while high water contents limits oxygen (O2) diffusion, creating anaerobic conditions that suppress aerobic microbial respiration (ref). Something here about modelling?

Forest disturbances have the potential to affect respiration rates and hence carbon cycling by modifying the microclimate as well as the decomposer communities (ref). Removal of trees by forest logging increases the exposure of soil to temperature fluctuations and alter the soil water content either by decreasing water through soil evaporation or by increasing it due to a reduction of the transpiration of water by trees (Zhang et al 2018). The degree of the response of soil is dependent on the severity of the forest disturbance. Thus, more severe disturbances such as clear-cuts seems to increase soil moisture, enhance heat and frost damage (Kovaks et al 2019; Zhang et al 2018) and reduce respiration rates (Yang et el 2022; Zhang et al 2018). Less severe disturbances such as thinning, have shown to increase respiration in broadleaf forests in a meta-analysis (Zhang et al 2018). Autotrophic respiration is very sensitive to the decrease in photosynthetic activity due to the loss of trees (Levy-Varon et al. 2012, 2014;Moore et al. 2013; Borkhuu et al. 2015). Instead, increases in heterotrophic respiration due to the increases of saprotrophic activity after the disturabcne

Negative effects of forest disturbances on soil could be buffered by the addition of forest residues on soil (i.e., slash addition), which can improve the microclimatic conditions, nutrient fertility and biodiversity (Pérez-Izquierdo et al. 2025). There is a lack of knowledge about the response of soil CO2 fluxes to the microclimatic conditions (i.e. soil temperature and moisture) under different forest disturbances and if this response can be modulated by effective post-disturbance management strategies. Understanding the sensitivity of respiration to microclimate under different forest disturbances is crucial in a context of climate-change-mediated increased disturbances.

Here, we hypothesized that:
(1) logging of trees will decrease the respiration rates because of the strong decrease in vegetation activity and microbiota associated. This decrease will be compensated by addition of slash which provides labile organic matter, increasing the heterotrophic respiration. 
(2) soil respiration would be more dependent on microclimatic variables (soil temperature and moisture), in the management disturbance treatments where the vegetation was removed and where the litter layer (slash) was removed, by depletion of labile organic matter. Likewise, the sensitivity of respiration to microclimate will decrease as the time passes since the disturbance because of new fresh C input from the growth of understory vegetation
(3) the sensitivity of soil respiration to microclimatic conditions will also depend on the type of climate/forest, being the Mediterranean the less sensitive as the microbial communities will be better adapted to the strong seasonal changes in water availability.

To test these hypotheses, we have recorded data on soil respiration and microclimate every month in three replicated Quercus forests across Europe subjected to different forest management strategies during three years. We have analysed the sensitivity of soil respiration to microclimate (i.e., soil temperature and soil moisture) by constructing a mechanistic parametric model and compared to a non-parametric model (i.e., random forest). We have also explored the main drivers of variation in Q10 across different forests and forest managements disturbances.

@@

## Study sites
The study sites are located in Gamiz, Northern Spain (42°49'2.32"N; 2°37'10.67"W), Saint-Christol-d’Albion, South of France (xxx) and Dumbravita, East Romania (42°49'2.32"N; 2°37'10.67"W). According to Köppen-Geiger climatic classification, the climate in Gamiz is Oceanic, with cold and rainy winters and warm summers. In St. Christol the climate is Mediterranean, with mild winters and warm and dry summers. In Dumbravita, the climate is humid continental with warm to hot summers, and cold and snowy winters with precipitation usually distributed throughout the year (Table 1). In the three study sites, the dominant vegetation are deciduous species of Quercus. Specifically, Quercus faginea Lam. in Gamiz, Quercus pubescens Willd. in St. Christol and Quercus robur L. in Dumbravita (Table 1). In St.Christol and Dumbravita the stands are Quercus monodominant, while in Gamiz, Quercus faginea appears together with scattered Acer campestris L. and Fraxinus excelsior L., and an understory dominated by Crataegus monogyna Jacq., Juniperus communis L. and Viburnum lantana L. More characteristics of the study sites are found in Table 1.
The experimental sites have an approximate total area of 1.5 ha. In each site, three stands of 0.12 ha each and replicated in four blocks were delimited. The stands corresponded to a control and two canopy disturbance treatments, i.e., 50% thinning and clear-cut (Fig. S1). The stands from disturbed treatments (thinning and clear-cut) were in turn divided in two: on half of the stands the slash resulting from the logging (leaves and twigs) was crashed and added on top of the soil and on the other half, the woody debris resulting from logging was removed. This made a total of 5 treatments (control, clear-cut with and without slash, thinning with and without slash) per block (Fig. S1). In each of the 5 treatments, two plots were established per block making a total of 40 plots (4 blocks x 5 treatments x 2 replicates per block; Fig S1). In turn, in each plot a transect of 2m length was established for monitoring (Fig. S1). Both the canopy disturbance and the slash addition treatments together with the fencing of the entire area were carried out in October 2021 in Gamiz, April 2022 in St. Christol and March 2022 in Dumbravita.

## Soil respiration 
To measure soil CO2 efflux (i.e., soil respiration), three rings of 10 cm diameter and 8 cm high were put along the 2 m transect in each of the monitoring plots (Fig. S1). Respiration was measured by a mobile infrared gas analyser (EGM-5, PP Systems, Hitchin, UK) and an attached soil respiration chamber (SRC-1). Soil CO2 efflux (g CO2 m-2 h-1) was calculated as linear increase in CO2 concentrations within the chamber headspace. The total volume used for calculation was the volume of the chamber plus the volume of the specific ring. Measurements were done monthly in the 120 collars in each site. Measurement in Gamiz started in November 2021, in St. Christol in February 2023 and in Dumbravita in May 2022.

## Forest stand characteristics
Within each of the control and thinned stands (n=8), we recorded the number of trees to calculate tree density (No. tree per ha) and measured tree diameter at breast height (DBH; 1.3 m) to calculate basal area.

@@

## Soil physical-chemical and enzymatic analyses
Chemical analyses and enzymes were measured on 5 soil cores of 4 cm diameter separated 0.5 m to a depth of 10 cm along the 2m monitoring transect. The five soil cores were mixed into a composite sample per plot. The composite sample was sieved through a 5-mm mesh sieve in the field, stored at -20°C and then freeze-dried prior to laboratory analyses. The pH was measured in distilled water 1:10 (w/v). The carbon (C) and nitrogen (N) content of the soil was measured using an elemental analyser: C content was measured by sulfochromic oxidation (ISO 14235), and N content was estimated by sulfuric acid mineralization with the addition of selenium and sodium sulphate and conversion to ammonium ions (ISO 11261), which were measured with the Segmented Flow Analyzer (SFA) Skalar. Available phosphorus (P) was determined after extraction with malachite green according to the protocol by Ohno & Zibilske (1991). Bulk density was measured in each of the plots by collecting 3 soil cores of 6 cm diameter and 10 cm depth around the 2m transect that were mixed into a composite sample. Soil bulk density was determined by dividing the dry mass of the composite sample by the volume of the 3 cores, after removing the stones. Soil C stocks (g m2) were calculated by multiplying C concentration by bulk density and sampling depth.
The activity of extracellular enzymes involved in SOM decomposition was measured through enzyme assays developed by Štursová & Baldrian (2011). Briefly, the activity of cellobiohydrolase, β-galactosidase, β-glucosidase, β-xylosidase, lipase, chitinase, α-glucosidase and acid-phosphatase  were measured at pH 5.0 in 1:100 (w/v) sample slurries using methylumbelliferol-based substrates, with an excitation wavelength of 355 nm and an emission wavelength of 460 nm. Product quantification was based on standard curves with a range of 4-methylumbelliferol concentrations in the same sample slurry.

@@

We included in this model tree density in plants ha$^{-1}$ ($trees$), average diameter at breast height ($DBH$), basal area ($BA$),  pH, bulk density ($BD$), phosophorous ($P$), total nitrogen ($N}$), carbon stocks ($C_{stocks$), fungal biomass ($biom_{fungi}$), bacterial biomass ($biom_{bact}$), actinomycetes biomass ($biom_{act}$), gram+ bacterial biomass ($biom_{gram+}$), gram- bacterial biomass ($biom_{gram-}$), total biomass ($biom_{tot}$), fungi:bacteria ratio, soil temperature ($T_{soil}$), soil moisture  ($M_{soil}$) and month of the year.
This model was tested only on a subset of the dataset (only the sites in Spain), for which all the additional variables were available.
The aim of this model was to test the predictive potential of any of these additional variables, in order to to explore eventual correlations and possible related processes. The choice of a random forest model over a more conventional linear model is motivated by the capacity of the former to consider more complex relationships and interactions, on top of the convenience of being able to assess quantitatively the predictive potential of each variable by estimating relative variable importance.

@@

## Characterization of study sites
Tree density in France was very high compared to Spain and especially to Romania where the tree density was very low (Figure 1). In contrast, the DBH was the lowest in France and the highest in Romania (Figure 1). Spain showed intermediate values of both tree density and DBH (Figure 1). For soil variables, in the untouched forest (control), the bulk density was the highest in Spain and the lowest in France. Spain had the highest pH while France and Romania had similar ranges of pH. The highest significant levels of C and N were found in Spain while the lowest C content was found in Romania. France in turn presented doubled C:N rates than the other two sites (Figure 1). Total microbial biomass, fungal biomass and bacterial biomass peaked in France, followed by Spain and it was the lowest in Romania. Similarly, in general, activities of most enzymes were higher in France compared with Spain and Romania (Figure 2).

![Figure1. Boxplots of stand characteristics and soil physical-chemical properties per disturbance treatment and study site. Red shades = France, green shades = Romania, blue shades = Spain](../Figures/edaphic_boxplots.png){width=80%}

## Effects of management disturbance treatments on soil physical-chemical and biological properties
The disturbance treatments barely affected the soil properties in Romania (Figure 1). In France, the bulk density increased as the intensity of the disturbance did (i.e., from thinning with slash to clear-cut no slash) and in Spain it significantly increased in all the disturbance treatments compared with control (Figure 1). The pH only significantly increased in both clear-cut treatments in Spain. Moreover, in Spain the N content decreased in the thinning and clear-cut without slash (Figure 1). In France, compared with controls, C content increased in all the disturbance treatment, but in the clear-cut without slash that decreased (Figure 1). In Spain, C content only decreased in the thinning-without slash treatment (Figure 1). 
Similar to soil properties, microbial biomass, fungal biomass and bacterial biomass did not change with the disturbance treatment in Romania. In Spain, these variables significantly decreased in the disturbance treatments where the slash was removed and in France only in the clear-cut treatment where the slash was removed (Figure 2). The disturbance treatments barely affected enzymatic activities in Romania and Spain, but generally decreased in clear-cut where slash was removed in France (Figure 2).

![Figure2. Boxplots of soil microbial biomass and enzymatic activities per disturbance treatment and study site. Red shades = France, green shades = Romania, blue shades = Spain](../Figures/enzymes_boxplots.png){width=80%}

@@

As expected, in Spain the temperature variation was the lowest compared to the other two sites (Figure 3a). In general, control plots showed the lowest thermal amplitude in all sites (i.e., with higher temperatures in winter and lower in summer). In Spain, the soil temperature increased as the intensity of the tree removal, mainly in summer, but the addition of slash buffered these increases. In France, the main effect on soil temperature was the tree removal, especially in summer and this was independent on the addition of slash. Soil temperature was not affected by disturbance and post-disturbance treatments in Romania. The observed effects of the disturbance and post-disturbance treatments on soil temperature were decreasing with time.
Soil moisture fluctuated inversely with temperature. In Spain, soil moisture was higher in the clear-cut with and without slash addition, except in the driest periods (from April to August). Similarly, in Romania, clear-cut increased soil moisture independently of slash addition. Contrarily, in France, soil moisture tended to decrease in the clear-cut treatment with and without slash (Figure 3b). 
Both in Spain and France, removal of slash decreased respiration rates, independently of tree cutting intensity, and thinning coupled with slash addition kept the respiration rates as the same level as controls. In Romania, effects of the disturbances were lower, however clear-cut increased respiration rates, especially when slash was added to the soil. At the end of the monitoring period, three years for Spain and two years for France and Romania after disturbance, respiration rates increased in the clear-cut with slash addition in the three study sites (Figure 3c).

@@

![Figure 5. Fit of the parametric model,  constant `E_0`](../Figures/fit_model_validation.png){width=60%}

## Temperature sensitivity (Q10) and parametric model posteriors 
Most of the between treatment and between sites variance was explained by the activation energy parameters of the model.
The temperature sensitivity (Q10) values of SOM mineralization were consistently higher in Romania (2.9 in controls), followed by France (2.7) and Spain (2.5) (Figure S4), as well as the activation energy (Ea) parameter wiich as expected followed the same pattern as Q10 (Figure 6a, Figure S4). In general, the activation energy and Q10 values decreased with slash addition due to higher SOM quality compared with treatments in which slash was removed. This was especially evident in Spain for both thinning and clear-cut and in the clear treatment in France. The activation energy and Q10 in the treatments where slash was added did not differ from controls (Figure 6a, Figure S4). These results were consistent when the activation energy was analysed as a function of time since disturbance and the activation energy systematically decreased in all treatments and all countries with time since disturbance (Figure 7). Unaccounted processes related to photosynthetic activity did not seem to have a substantial impact on the variance of respiration, since posterior values for the amplitude parameter of the sine function were close to zero (Figure 6b, c). However, the amplitude of the sine-function tended to be higher in France (Figure 6c). The parameters of soil moisture function (a and b) did not explain a big part of the treatment and site related variance either (Figure 6d, e), since posteriors were substantially indistinguishable.  
The scaling parameter $A$ (Figure S5) still seems needed to explain part of the variance in the measured respiration, even after normalizing the respiration for initial C stocks.

@@

Hydrologic leaching of dissolved and particulate carbon compounds can also be important outputs of C from the system (Davinson and Jansen 2006), that could explained the low carbon content in Romania (sandy soil with low retention capacity).

The scaling parameter $A$ was still needed to explain some of the variance in respiration measured at the level of single experimental plots even after normalization by SOC stocks. This is likely because of the high variability in SOC content of soils, which, particularly in forest soils, can be affected by local processes at a very small spatial scale. For example small depressions and local micromorphology, root concentration, localized events such as branch or tree falls, and so on, can substantially affect SOC at a metric or submetric scale. This could easily cause discrepancies between the SOC stocks at the site of each respiration measuring collar and the SOC stocks measured by pooling at the level of each experimental plot. Since respiration depends first of all on the amoutn of substrate, local stocks can heavily affect it. By allowing the $A$ parameter to vary at the level of the experimental plot we could allow for the needed freedom for our parametric models to account for this effect (at the cost of an increased variance of the posterior, but reducing eventual biases).

@@

## Forest management effect on activation energy and sensitivity to temperature
Generally, our results confirmed the hypothesis 1, since slash addition increased the respiration rates compared to the treatments where the slash was removed. Addition of slash created an artificial litter layer that buffered the effects over soil microclimatic conditions, and provided new habitat and food availability for soil organisms (Pérez-Izquierdo et al. in prep). These results indicate that increases in heterotrophic respiration due to an increment in organic matter availability by slash addition compensated for the decreases in autotrophic respiration and root-associated microbiota due to tree logging and soil disturbance (Borkhuu et al. 2015). Moreover, the strength of the response was also dependent on the forest, being more evident in Spain and France where the tree density was much higher than Romania. In Romania such a low density of trees together with a dense understory of grasses in all treatments might explain the lack of response and indicate high degree of resilience to perturbations.

Our results also confirmed the hypothesis 2, since addition of slash in general decreased the activation energy of SOM decomposition, particularly when in a clearcut (which is already adding fresh inputs to the soil with residuals like leaves and branches falling to the ground as well as roots that start to decompose). We argue that this result is probably related to a reduction in chemical recalcitrance due to the increased amount of fresh C inputs, supporting the carbon quality-temperature hypothesis stating that the biochemically recalcitrant C pool is more sensitive to changes in temperature than the labile C pool [@Bosatta.1999], confirmed by other direct measurements [@Lefevre.2014]. A more recalcitrant substrate likely implies more biochemical steps for its decomposition, each step subjected to the same thermodynamic constraints. More steps in the overall oxydative reaction leading also implies a higher chance of one being more limited than the others due to the enzyme characteristics.

>>>>

# Introduction
The soil is one of the largest terrestrial reservoirs of carbon (C), containing approximately three times more C than the atmosphere or the vegetation on Earth (IPCC, 2021). Soil C is released to the atmosphere as carbon dioxide (CO2) through decomposition of organic matter by microbes (heterotrophic respiration) and root respiration (autotrophic respiration). This process (i.e., soil respiration) constitutes the main C efflux into the atmosphere (ref). Under the new climate change scenarios increases of CO2 efflux from soils are expected (Schlesinger and Andrews, 2000), this can imply a positive feedback to future warming if stored C is mobilized and C input to soil does not increase (Lamberty&Thomson, 2010). Because soil respiration is highly sensitive to soil temperature and soil moisture (Nissan et al. 2022), understanding the responses of soil respiration to these microclimatic conditions, which are affected by global change factors as climate change and forest management, is critical for predicting future atmospheric CO2 concentrations. Temperature affects respiration rates positively (Lloyd and Taylor, 1994), by increasing the metabolic rates of soil microorganisms and enzyme efficiency (ref). However, the relationship with soil moisture is non-linear. Low water content limit nutrient diffusion and enzymes and exert drought-stress on microbes while high water contents limits oxygen (O2) diffusion, creating anaerobic conditions that suppress aerobic microbial respiration (ref). Something here about modelling?

Forest disturbances have the potential to affect respiration rates and hence C cycling by modifying the microclimate, nutrients as well as the decomposer communities (ref). Removal of trees by forest logging increases the exposure of soil to temperature fluctuations and alter the soil water content either by decreasing water through soil evaporation or by increasing it due to a reduction of the transpiration of water by trees (Zhang et al 2018). The degree of the response of soil is dependent on the severity of the forest disturbance. Thus, more severe disturbances such as clear-cuts seems to increase soil moisture, enhance heat and frost damage (Kovaks et al 2019; Zhang et al 2018) and reduce respiration rates (Yang et el 2022; Zhang et al 2018). Less severe disturbances such as thinning, have shown to increase respiration in broadleaf forests in a meta-analysis (Zhang et al 2018). After a disturbance, imbalances between autotrophic and heterotrophic respiration might happen (ref). The autotrophic respiration is very sensitive to the decrease in photosynthetic activity due to the loss of trees (Levy-Varon et al. 2012, 2014;Moore et al. 2013; Borkhuu et al. 2015). Instead, increases in heterotrophic respiration due to the increases in saprotrophic activity after the disturbance are normal (ref).

Negative effects of forest disturbances on soil could be buffered by the addition of forest residues on soil (i.e., slash addition), which can improve the microclimatic conditions, nutrient fertility and biodiversity (Pérez-Izquierdo et al. 2025). Moreover, addition of slash in turn might decrease the temperature sensitivity of soil respiration, making it less vulnerable to climate-induced temperature raise. There is a lack of knowledge about the response of soil CO2 fluxes to the microclimatic conditions (i.e. soil temperature and moisture) under different forest disturbances and if this response can be modulated by effective post-disturbance management strategies. Understanding the sensitivity of respiration to microclimate under different forest disturbances is crucial in a context of climate-change-mediated increased disturbances.

Here, we hypothesized that:
(1) canopy disturbances will decrease the respiration rates because of the strong decrease in vegetation activity and microbiota associated. This decrease will be compensated by addition of slash which provides labile organic matter, increasing the heterotrophic respiration. 
(2) soil respiration would be more dependent on microclimatic variables (soil temperature and moisture), in the disturbance treatments where the vegetation was removed and where the litter layer (slash) was removed, by depletion of labile organic matter. Likewise, the sensitivity of respiration to microclimate will decrease as the time passes since the disturbance because of new fresh C input from the growth of understory vegetation
(3) the sensitivity of soil respiration to microclimatic conditions will also depend on the type of climate/forest, being the Mediterranean the less sensitive as the microbial communities will be better adapted to the strong seasonal changes in water availability.

To test these hypotheses, we have recorded data on soil respiration and microclimate every month in three replicated Quercus forests across Europe subjected to different intensities of forest canopy disturbances during three years. We have analysed the sensitivity of soil respiration to microclimate (i.e., soil temperature and soil moisture) by constructing a mechanistic parametric model and compared to a non-parametric model (i.e., random forest). We have also explored the main drivers of variation in Q10 across different forests and forest canopy disturbances.

@@

## Study sites
The study sites are located in Gamiz, Northern Spain (42°49'2.32"N; 2°37'10.67"W), Saint-Christol-d’Albion, South of France (44°02'58.0"N 5°32'35.8"E) and Dumbravita, East Romania (42°49'2.32"N; 2°37'10.67"W). According to Köppen-Geiger climatic classification, the climate in Gamiz is Oceanic, with cold and rainy winters and warm summers. In St. Christol the climate is Mediterranean, with mild winters and warm and dry summers. In Dumbravita, the climate is humid continental with warm to hot summers, and cold and snowy winters with precipitation usually distributed throughout the year (Table 1). In the three study sites, the dominant vegetation are deciduous species of Quercus. Specifically, Quercus faginea Lam. in Gamiz, Quercus pubescens Willd. in St. Christol and Quercus robur L. in Dumbravita (Table 1). In St.Christol and Dumbravita the stands are Quercus monodominant, while in Gamiz, Quercus faginea appears together with scattered Acer campestris L. and Fraxinus excelsior L., and an understory dominated by Crataegus monogyna Jacq., Juniperus communis L. and Viburnum lantana L. More characteristics of the study sites are found in Table 1.
The experimental sites have an approximate total area of 1.5 ha. In each site, three stands of 0.12 ha each and replicated in four blocks were delimited. The stands corresponded to a control and two canopy disturbance treatments, i.e., 50% canopy removal (CD50) and 100 % canopy removal (CD100)  (Fig. S1). The stands from disturbed treatments (CD50 and CD100) were in turn divided in two: on half of the stands the slash resulting from the logging (leaves and twigs) was crashed and added on top of the soil and on the other half, the woody debris resulting from logging was removed. This made a total of 5 treatments (control, CD50 with and without slash, CD100 with and without slash) per block (Fig. S1). In each of the 5 treatments, two plots were established making a total of 40 plots (4 blocks x 5 treatments x 2 replicates; Fig S1). In turn, in each plot a transect of 2m length was established for monitoring (Fig. S1). Both the canopy disturbance and the slash addition treatments together with the fencing of the entire area were carried out in October 2021 in Gamiz, April 2022 in St. Christol and March 2022 in Dumbravita.

## Soil respiration 
To measure soil CO2 efflux (i.e., soil respiration), three rings of 10 cm diameter and approximately 8 cm high were put along the 2 m transect in each of the monitoring plots (Fig. S1). Respiration was measured by a mobile infrared gas analyser (EGM-5, PP Systems, Hitchin, UK) and an attached soil respiration chamber (SRC-1). Soil CO2 efflux (g CO2 m-2 h-1) was calculated as linear increase in CO2 concentrations within the chamber headspace. The total volume used for calculation was the volume of the chamber plus the volume of the specific ring. Measurements were done monthly in the 120 collars in each site. Measurement in Gamiz started in November 2021, in St. Christol in February 2023 and in Dumbravita in May 2022.

## Forest stand characteristics
Within each of the control and CD50 stands (n=8), we recorded the number of trees to calculate tree density (No. tree per ha) and measured tree diameter at breast height (DBH; 1.3 m) to calculate basal area.

@@

## Soil physical-chemical and enzymatic analyses
Chemical analyses and enzymes were measured on 5 soil cores of 4 cm diameter separated 0.5 m to a depth of 10 cm along the 2m monitoring transect. The five soil cores were mixed into a composite sample per plot. The composite sample was sieved through a 5-mm mesh sieve in the field, stored at -20°C and then freeze-dried prior to laboratory analyses. The pH was measured in distilled water 1:10 (w/v). The C and nitrogen (N) content of the soil was measured using an elemental analyser: C content was measured by sulfochromic oxidation (ISO 14235), and N content was estimated by sulfuric acid mineralization with the addition of selenium and sodium sulphate and conversion to ammonium ions (ISO 11261), which were measured with the Segmented Flow Analyzer (SFA) Skalar. Available phosphorus (P) was determined after extraction with malachite green according to the protocol by Ohno & Zibilske (1991). Bulk density was measured in each of the plots by collecting 3 soil cores of 6 cm diameter and 10 cm depth around the 2m transect that were mixed into a composite sample. Soil bulk density was determined by dividing the dry mass of the composite sample by the volume of the 3 cores, after removing the stones. Soil C stocks (g m2) were calculated by multiplying C concentration by bulk density and sampling depth.
The activity of extracellular enzymes involved in SOM decomposition was measured through enzyme assays developed by Štursová & Baldrian (2011). Briefly, the activity of cellobiohydrolase, β-galactosidase, β-glucosidase, β-xylosidase, lipase, chitinase, α-glucosidase and acid-phosphatase  were measured at pH 5.0 in 1:100 (w/v) sample slurries using methylumbelliferol-based substrates, with an excitation wavelength of 355 nm and an emission wavelength of 460 nm. Product quantification was based on standard curves with a range of 4-methylumbelliferol concentrations in the same sample slurry.

@@

We included in this model tree density in plants ha$^{-1}$ ($trees$), average diameter at breast height ($DBH$), basal area ($BA$),  pH, bulk density ($BD$), phosophorous ($P$), total nitrogen ($N}$), C stocks ($C_{stocks$), fungal biomass ($biom_{fungi}$), bacterial biomass ($biom_{bact}$), actinomycetes biomass ($biom_{act}$), gram+ bacterial biomass ($biom_{gram+}$), gram- bacterial biomass ($biom_{gram-}$), total biomass ($biom_{tot}$), fungi:bacteria ratio, soil temperature ($T_{soil}$), soil moisture  ($M_{soil}$) and month of the year.
This model was tested only on a subset of the dataset (only the sites in Spain), for which all the additional variables were available.
The aim of this model was to test the predictive potential of any of these additional variables, in order to to explore eventual correlations and possible related processes. The choice of a random forest model over a more conventional linear model is motivated by the capacity of the former to consider more complex relationships and interactions, on top of the convenience of being able to assess quantitatively the predictive potential of each variable by estimating relative variable importance.

@@

## Characterization of study sites
Tree density in France was very high compared to Spain and especially to Romania where the tree density was very low (Figure 1). In contrast, the DBH was the lowest in France and the highest in Romania (Figure 1). Spain showed intermediate values of both tree density and DBH (Figure 1). For soil variables, in the untouched forest (i.e., control), the bulk density was the highest in Spain and the lowest in France. Spain had the highest pH while France and Romania had similar ranges of pH. The highest significant levels of C and N were found in Spain while the lowest C content was found in Romania. France in turn presented doubled C:N rates than the other two sites (Figure 1). Total microbial biomass, fungal biomass and bacterial biomass peaked in France, followed by Spain and it was the lowest in Romania. Similarly, in general, activities of most enzymes were higher in France compared with Spain and Romania (Figure 2).

![Figure1. Boxplots of stand characteristics and soil physical-chemical properties per disturbance treatment and study site. Red shades = France, green shades = Romania, blue shades = Spain. CD50 = 50% canopy disturbance; CD100 = 100% canopy disturbance; +S = with slash; -S = without slash.](../Figures/edaphic_boxplots.png){width=80%}

## Effects of management disturbance treatments on soil physical-chemical and biological properties
The disturbance treatments barely affected the soil properties in Romania (Figure 1). In France, the bulk density increased as the intensity of the disturbance did (i.e., from CD50 with slash to CD100 no slash) and in Spain it significantly increased in all the disturbance treatments compared with control (Figure 1). The pH only significantly increased in both CD100 treatments in Spain. Moreover, in Spain the N content decreased in the CD50 and CD100 without slash (Figure 1). In France, compared with controls, C content increased in all the disturbance treatment, but in the CD100 without slash that decreased (Figure 1). In Spain, C content only decreased in the CD50 without slash treatment (Figure 1). 
Similar to soil properties, microbial biomass, fungal biomass and bacterial biomass did not change with the disturbance treatments in Romania. In Spain, these variables significantly decreased in the disturbance treatments where the slash was removed and in France only in the CD100 treatment where the slash was removed (Figure 2). The disturbance treatments barely affected enzymatic activities in Romania and Spain, but generally decreased in the CD100 where slash was removed in France (Figure 2).

![Figure2. Boxplots of soil microbial biomass and enzymatic activities per disturbance treatment and study site. Red shades = France, green shades = Romania, blue shades = Spain. CD50 = 50% canopy disturbance; CD100 = 100% canopy disturbance; +S = with slash; -S = without slash.](../Figures/enzymes_boxplots.png){width=80%}

@@

As expected, in Spain the temperature variation was the lowest compared to the other two sites (Figure 3a). In general, control plots showed the lowest thermal amplitude in all sites (i.e., with higher temperatures in winter and lower in summer). In Spain, the soil temperature increased as the intensity of the canopy disturbance, mainly in summer, but the addition of slash buffered these increases. In France, the main effect on soil temperature was the canopy disturbance, especially in summer and this was independent on the addition of slash. Soil temperature was not affected by disturbance and post-disturbance treatments in Romania. The observed effects of the disturbance and post-disturbance treatments on soil temperature were decreasing with time.
Soil moisture fluctuated inversely with temperature. In Spain, soil moisture was higher in the CD100 with and without slash addition, except in the driest periods (from April to August). Similarly, in Romania, CD100 increased soil moisture independently of slash addition. Contrarily, in France, soil moisture tended to decrease in the CD100 treatment with and without slash (Figure 3b). 
Both in Spain and France, removal of slash decreased respiration rates, independently of canopy disturbance intensity, and CD50 coupled with slash addition kept the respiration rates as the same level as controls. In Romania, effects of the disturbances were lower, however CD100 increased respiration rates, especially when slash was added to the soil. At the end of the monitoring period, three years for Spain and two years for France and Romania after disturbance, respiration rates increased in the CD100 with slash addition in the three study sites (Figure 3c).

@@

![Figure 5. Fit of the parametric model,  constant `E_0`. Red shades = France, green shades = Romania, blue shades = Spain. CD50 = 50% canopy disturbance; CD100 = 100% canopy disturbance; +S = with slash; -S = without slash.](../Figures/fit_model_validation.png){width=60%}

## Temperature sensitivity (Q10) and parametric model posteriors 
Most of the between treatment and between sites variance was explained by the activation energy parameters of the model.
The temperature sensitivity (Q10) values of SOM mineralization were consistently higher in Romania (2.9 in controls), followed by France (2.7) and Spain (2.5) (Figure S4), as well as the activation energy (Ea) parameter which as expected followed the same pattern as Q10 (Figure 6a, Figure S4). In general, the activation energy and Q10 values decreased with slash addition compared with treatments in which slash was removed. This was especially evident in Spain for both CD50 and CD100 and in the CD100 treatment in France. The activation energy and Q10 in the treatments where slash was added did not differ from controls (Figure 6a, Figure S4). These results were consistent when the activation energy was analysed as a function of time since disturbance, and the activation energy systematically decreased in all disturbance treatments and all countries with time since disturbance (Figure 7). Unaccounted processes related to photosynthetic activity did not seem to have a substantial impact on the variance of respiration, since posterior values for the amplitude parameter of the sine function were close to zero (Figure 6b, c). However, the amplitude of the sine-function tended to be higher in France (Figure 6c). The parameters of soil moisture function (a and b) did not explain a big part of the treatment and site related variance either (Figure 6d, e), since posteriors were substantially indistinguishable.  
The scaling parameter $A$ (Figure S5) still seems needed to explain part of the variance in the measured respiration, even after normalizing the respiration for initial C stocks.

@@

Hydrologic leaching of dissolved and particulate C compounds can also be important outputs of C from the system (Davinson and Jansen 2006), that could explained the low C content in Romania (sandy soil with low retention capacity).

The scaling parameter $A$ was still needed to explain some of the variance in respiration measured at the level of single experimental plots even after normalization by SOC stocks. This is likely because of the high variability in SOC content of soils, which, particularly in forest soils, can be affected by local processes at a very small spatial scale. For example, small depressions and local micromorphology, root concentration, localized events such as branch or tree falls, and so on, can substantially affect SOC at a metric or submetric scale. This could easily cause discrepancies between the SOC stocks at the site of each respiration measuring collar and the SOC stocks measured by pooling at the level of each experimental plot. Since respiration depends first of all on the amount of substrate, local stocks can heavily affect it. By allowing the $A$ parameter to vary at the level of the experimental plot we could allow for the needed freedom for our parametric models to account for this effect (at the cost of an increased variance of the posterior, but reducing eventual biases).

@@

## Forest management effect on activation energy and sensitivity to temperature
Generally, our results confirmed the hypothesis 1, since slash addition increased the respiration rates compared to the treatments where the slash was removed. Addition of slash created an artificial litter layer that buffered the effects over soil microclimatic conditions, and provided new habitat and food availability for soil organisms (Pérez-Izquierdo et al. in prep). These results indicate that increases in heterotrophic respiration due to an increment in organic matter availability by slash addition compensated for the decreases in autotrophic respiration and root-associated microbiota due to the canopy and soil disturbance (Borkhuu et al. 2015). Moreover, the strength of the response was also dependent on the forest, being more evident in Spain and France where the tree density was much higher than Romania. In Romania such a low density of trees together with a dense understory of grasses in all treatments might explain the lack of response and indicate high degree of resilience to perturbations.

Our results also confirmed the hypothesis 2, since addition of slash in general decreased the activation energy of SOM decomposition, particularly in CD100 (which is already adding fresh inputs to the soil with residuals like leaves and branches falling to the ground as well as roots and mycelium that start to decompose). We argue that this result is probably related to a reduction in chemical recalcitrance due to the increased amount of fresh C inputs, supporting the C quality-temperature hypothesis stating that the biochemically recalcitrant C pool is more sensitive to changes in temperature than the labile C pool [@Bosatta.1999], confirmed by other direct measurements [@Lefevre.2014]. A more recalcitrant substrate likely implies more biochemical steps for its decomposition, each step subjected to the same thermodynamic constraints. More steps in the overall oxydative reaction leading also implies a higher chance of one being more limited than the others due to the enzyme characteristics.
